Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-21 paper.qmd
Edit paper.qmd for clarity and consistency
Revised text for clarity and consistency throughout the paper, including updates to the narrative voice and the presentation of tools and frameworks.

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. We present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager, making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing our approach, we survey existing tools and
their limitations to contextualize our contribution.

### Existing approaches to environment management {#sec-existing}

@@

#### Package-level solutions: renv and alternatives

@@

#### System-level solutions: Docker

@@

### The workflow orchestration gap

@@

### A unified approach with Nix

@@

[Nix]{.pkg} provides strong cross-platform support across major operating systems. It runs natively on Linux 
(x86_64) with full functionality and optimal integration. On macOS (Intel and Apple Silicon), however, 
reproducibility is less reliable due to dependencies on impure system frameworks and Xcode toolchains outside
the Nix store. As a result, macOS users may need to update project pins after system or Xcode upgrades to 
restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
applications may need extra setup. Nix environments on WSL can be used interactively via VS Code or Positron 
for a smooth development experience.

@@

This call to `rix()` generates an environment that installs several
[R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
[fusen]{.pkg} from GitHub, and the Positron editor.

@@

Additionally, updating the full [R]{.proglang} package set on [Nix]{.pkg} daily
is impractical. While CRAN and Bioconductor update daily, the [R]{.proglang}
packages in `nixpkgs` are only updated with new R releases. This limitation is
due to [Nix]{.pkg}’s governance as a Linux distribution package manager.

Our fork allows us to circumvent these limitations. For example, we provide a
daily snapshot of CRAN. Each day, the [R]{.proglang} package set is updated and
committed to a dated branch using GitHub Actions. Users can select a specific
date with:

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. We
backported the fix to our fork, making RStudio available to users of our fork
earlier than upstream, as merging PRs in the official repository can take some
time.

We have backported fixes to our `nixpkgs` fork as far back as March 2019. The
process involves checking out a `nixpkgs` commit on the selected date, updating
the [R]{.proglang} package set using Posit CRAN and Bioconductor snapshots,
backporting fixes, and ensuring popular packages work on both x86-linux
(including WSL2) and aarch64-darwin (Apple Silicon). These changes are committed
to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all available dates
with `rix::available_dates()`.

@@

@@

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. The
package can then generate a visual representation of the pipeline's directed
acyclic graph:

@@

>>>>

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager, making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing our approach, I survey existing tools and
their limitations to contextualize our contribution.

@@

@@

@@

@@

@@

[Nix]{.pkg} provides strong cross-platform support across major operating systems. It runs natively on Linux 
(x86_64) with full functionality and optimal integration. On macOS (Intel and Apple Silicon), however, 
reproducibility is less reliable due to dependencies on impure system frameworks and Xcode toolchains outside
the Nix store. As a result, macOS users may need to update project pins after system or Xcode upgrades to 
restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
applications may need extra setup. [Nix]{.pkg} environments on WSL can be used interactively via VS Code or Positron 
for a smooth development experience. It is also important to note that [Nix]{.pkg} can be installed
within a [Docker]{.pkg} image to deploy and reproduce environments reliably in containerized workflows.

@@

This call to `rix()` generates a `default.nix` that builds a development shell
which ships [R]{.proglang} and [R]{.proglang} packages (including [AER]{.pkg} version 1.2-8), several TeXLive
packages for \LaTeX document authoring, development versions of [rix]{.pkg} and
[fusen]{.pkg} from GitHub, and the Positron editor.

@@

Additionally, updating the full [R]{.proglang} package set on [Nix]{.pkg} daily
is impractical. While CRAN and Bioconductor update daily, the [R]{.proglang}
packages in `nixpkgs` are only updated with new [R]{.proglang} releases. This limitation is
due to [Nix]{.pkg}’s governance as a Linux distribution package manager.

The `rstats-on-nix` fork allows us to circumvent these limitations. For example, it provides
daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
committed to a dated branch using GitHub Actions. Users can select a specific
date with:

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. We
backported the fix to our fork, making RStudio available to users of our fork
earlier than upstream, as merging PRs in the official repository can take some
time.

I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as March 2019. The
process involves checking out a `nixpkgs` commit on the selected date, updating
the [R]{.proglang} package set using Posit CRAN and Bioconductor snapshots,
backporting fixes, and ensuring popular packages work on both x86-linux
(including WSL2) and aarch64-darwin (Apple Silicon). These changes are committed
to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all available dates
with `rix::available_dates()`.

@@

The fork also allows me to catch issues (such as packages’ builds breaking) early
on, and prepare fixes that can then be contributed upstream.

@@

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow.
Data flows from derivation to derivation by being serialized into the 
efficient and language-agnostic  Arrow format using the pair of `encoder/decoder` 
functions. Other universal formats, such as `csv` or `json` could have been used.

The package can then generate a visual representation of the pipeline's directed
acyclic graph:

@@

While [rix]{.pkg} and [rixpress]{.pkg} provide significant steps forward, it is 
important to acknowledge current limitations. Both packages are relatively recent. 
[rixpress]{.pkg}, in particular, requires further development, especially to enhance
the ease of debugging complex pipelines. Its visualization capabilities are also still
fairly limited; for instance, they do not currently indicate outdated derivations when
a user modifies code that would necessitate a rebuild. These areas represent important 
avenues for future work to further mature the framework and improve the user experience.
